Concurrent Programming in Harmony

### 2021-12-16 book/paper.tex
Latest
Still working on debugging chapter

Looking further at the stack traces, we can see that
T3 was in the process of executing
\texttt{release}(?$q$.\texttt{lock}) within
\texttt{get}(?$q$).  Looking in the top green row,
we see that the value of $q$ is
\{ \texttt{.lock}: \texttt{True}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \}.
T1 is currently executing
\texttt{acquire}(\texttt{?alloc}.\textit{pool}[0].\texttt{lock})
within \texttt{put}(?$q$, 2), but 
\texttt{alloc}.\textit{pool}[0] does not exist.
The corresponding line of Harmony code is
\textbf{atomically when not} !\textit{binsema}
in line 25 of the \texttt{sync} module.

So how did we get there?  In the top we can see that the order of
events was the following:
\begin{enumerate}
\item[0:] initialization completed, leaving the value of $q$ at
\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \};
\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
\texttt{put}(1) (see the output column for that clue: the thread printed that).
We can see that $q$.\texttt{next} now points to \texttt{alloc}.\textit{pool}[0],
which the thread must have allocated.  The contents is 
\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: 1 \}, as expected;
\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
\texttt{put}(?$q$, 2).  We can see it got as far as putting 2 on the queue;
\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
\texttt{get}($?q$).  We can also see that it freed \textit{pool}[0].
We know that it is in the process of releasing $q$.\texttt{lock};
\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
no longer exists because T3 just freed it.
\end{enumerate}

Clearly there was a race in which T2 was trying to lock
?\textit{pool}[0].\texttt{lock} (which contained the node with the value 1)
while T3 was freeing that very same node, and T2 lost the race.

>>>>

Looking further at the stack traces, we can see that
T3 was in the process of executing
\texttt{release}(?$q$.\texttt{lock}) within
\texttt{get}(?$q$).
% Looking in the top green row, we see that the value of $q$ is
% \{ \texttt{.lock}: \texttt{True}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \}.
T1 is currently executing
\texttt{acquire}(\texttt{?alloc}.\textit{pool}[0].\texttt{lock})
within \texttt{put}(?$q$, 2), but 
\texttt{alloc}.\textit{pool}[0] does not exist.
The corresponding line of Harmony code is
\textbf{atomically when not} !\textit{binsema}
in line 25 of the \texttt{sync} module.

So how did we get there?  In the top we can see that the order of
events was the following:
\begin{enumerate}
\item[0:] initialization completed, with $q$ being
\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: \texttt{None} \};
\item[1:] thread T1 (\texttt{thread}(1)) ran and finished executing
\texttt{put}(1) (see the output column for that clue: the thread printed that).
We can see that $q$.\texttt{next} now points to \texttt{alloc}.\textit{pool}[0],
which the thread must have allocated.  The contents is 
\{ \texttt{.lock}: \texttt{False}, \texttt{.next}: \texttt{None}, \texttt{.value}: 1 \}, as expected;
\item[2:] thread T2 (\texttt{thread}(1)) started running, calling
\texttt{put}(?$q$, 2).  We can see it got as far as putting 2 on the queue,
but it is not yet done.  It is currently trying to acquire
\texttt{alloc}.\textit{pool}[0].\texttt{lock};
\item[3:] thread T3 (\texttt{thread}(1)) started running, calling
\texttt{get}($?q$).  We can also see that it freed \textit{pool}[0],
and is now releasing $q$.\texttt{lock};
\item[4:] thread T2 resumes and tries to access \textit{pool}[0], which
no longer exists (because T3 just freed it).
\end{enumerate}

Clearly there was a race in which T2 was trying to lock
?\textit{pool}[0].\texttt{lock} (which contained the node with the value 1)
while T3 was freeing that very same node, and T2 lost the race.
More precisely, T2 was executing \texttt{put}(?$q$, 2), when T3
preempted it with \texttt{get}(?$q$) and removed the node that T2
was trying to access.  But why did the locks not prevent this?

\index{reserve debugging}%
It is time to start stepping through the code that has been executed
before this happened.  This is sometimes known as
\emph{reverse debugging}.  In fact, Harmony allows you to step through
code forwards and backwards.  In this case, we first want to see what
T2 is doing.  You can click on its first (top-left) orange box to
time-travel to that part in the execution.  Now by hitting
$<$return$>$ repeatedly, we can quickly skip through the code.
T2 first calls \texttt{put}(?$q$, 1) and then allocates a new
node initialized with a lock.  Keep stepping until it executes
$\mathit{nq} = q$.  Hit $<$return$>$ once more and inspect the
state of T2 in the lower-right corner.  You can see that variable
\textit{nq} is initialized to~?$q$.  T2 then enters into the
\textbf{while} loop and tries to acquire
$\mathit{nq}$$\rightarrow$$\mathtt{lock}$.  This succeeds, and
next T2 executes \textbf{let} $n = \mathit{nq}$$\rightarrow$$\mathtt{next}$.
Now $n = ?\mathtt{alloc}.\mathit{alloc}[0]$, which is not \texttt{None}.
It then releases $\mathit{nq}$$\rightarrow$$\mathtt{lock}$
(\textit{nq} points to $q$).  It then sets \textit{nq} to $n$, which
is still $?\mathtt{alloc}.\mathit{alloc}[0]$.  Finally, it calls
$\mathtt{acquire}(?\mathit{nq}$$\rightarrow$$\mathtt{lock})$.
But before it can complete that operation, T3 runs next.

T3 chooses "get" and then goes on to invoke $\mathtt{get}(?q)$.
This first successfull acquires $q$$\rightarrow$$\texttt{lock}$.
It then finds out that $q$$\rightarrow$$\texttt{next}$ points to
$\texttt{alloc}.\textit{pool}[0]$.  It sets \textit{node} to
$\texttt{alloc}.\textit{pool}[0]$ as well and sets
$q$$\rightarrow$$\texttt{next}$ to 
$\mathtt{node}$$\rightarrow$$\texttt{next}$.

XXXXXXXX
